---
title: "Explore effort data - LME 42 - South East Australia"
output: 
  html_document:
    toc: yes
    toc_float: yes
editor_options: 
  chunk_output_type: console
---

This file explores the global fishing effort data that will be used by the Fish-MIP community to force regional and global marine ecosystem models. To understand how this forcing can be integrated into regional models, we take the South East Australia system (LME 42) as a case study and aim at developing a framework that can be applied across regions (with necessary adjustments). We chose LME 42 because three ecosystem models are available for this region: Atlantis, EwE and Mizer. As a first step, we compares the LME 42 extraction of the global effort data with effort data used to force the three regional models. Specifically, we:  

1. Check whether effort trends across these data match   
2. Explore how the global effort can be allocated across regional models' fleets     

Atlantis, EwE and Mizer differ in terms of area covered and fisheries considered, so some deviation in effort trends is expected:  

* Atlantis is the more comprehensive model, it covers LME 42 and 43 and all main Commonwealth fisheries 
* Mizer covers LME 42 and focuses on the commonwealth trawling and shark gillnet fishery  
* EwE models the Bass Strait (a small subset of LME42) and considers both State and Commonwealth fisheries  

These models can be considered nested in terms of:   

* Area covered Atlantis -> Mizer -> EwE  
* Fisheries and fleets considered: Atlantis -> EwE (or vice versa?) -> Mizer  
This is key when interpreting trends in effort and exploring how effort can be allocated to fleets

Effort is used differently across these models:  

* Atlantis and EwE are forced by fishing effort   
* Mizer is forced by either fishing mortality, or fish price and fisheries costs - effort is used only for validation aims.   

Effort units:  

* global data: days at sea X kW  
* Atlantis data: days at sea    
* EwE data: relative effort (days at sea also available now)  
* Mizer data: number of operations   

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

## Load libraries 

```{r libraries}

rm(list=ls())

# load libraries 
library("data.table")
library("ggplot2")
library("tidyverse")
library("patchwork")
library("gridExtra")
library("dplyr")
# library("networkD3") # PROBLEM WITH THIS - what is it for anyway??? 
library("tidyr")
# install.packages("patchwork")
# install.packages("networkD3")
# install.packages("igraph")

# # for stat analysis # WARNING UNABLE TO INSTALL PACKAGES ON R 
# install.packages("MuMIn")
# install.packages("RColorBrewer")
# install.packages("GGally")
library("MuMIn")
library("GGally")

select<-dplyr::select
summarise<-dplyr::summarise

```

## Global effort data 

### Load and explore data

```{r effort global}

# # OLD 
# effort<-read.csv("/Users/camillan/Dropbox/DBPM_fishing_extension/Yannick_data/TotalEffortby_FGroup_LME_Sector.csv")
# effort<-effort[2:nrow(effort), 2:ncol(effort)]

# NEW 
# effort<-read.csv("/Users/camillan/Dropbox/DBPM_fishing_extension/Yannick_data/DataYannick/TotalAggregatedFGroupLME.csv")
# effort<-effort[2:nrow(effort), 2:ncol(effort)]
# head(effort)

# New aggregation including gear 
yannick_dir <- "/rd/gem/private/users/yannickr"
effort<-read_csv(file.path(yannick_dir, "effort_histsoc_1961_2010.csv")) # OR all_effort_aggregated.csv as the first is still to be checked 
# effort<-effort[2:nrow(effort), 2:ncol(effort)]
head(effort)

# WARNING  - try fread from data.table instead of read_csv, then lazy_dt(effort) from dtplyr, then can apply dplyr sintax to data.table object which is much faster...  

# compare mapped effort with unmapped - what are the common columns in effort mapped and effort unmapped?
# load effort un-mapped
effort_unmapped<-read_csv(file.path(yannick_dir, "Final_DataStudyFAO_AllGears_V5.csv")) # WARNING - was originally this Final_DataStudyFAO_V5.csv in local
head(effort_unmapped)

# country, sector, and gear are the common columns, where country is the fishing country and not the eez
# View(unique(effort_unmapped[,c(3,8)]))

effort_mapped<-effort %>% 
  filter(SAUP == 36)  %>% # SAUP 36 = Au fishing country
  group_by(Year, Gear) %>% 
  summarise(effort = sum(NomActive, na.rm = TRUE)) %>% 
  ungroup() %>%
  mutate(origin = "mapped")

# # load effort unmapped
# # is the pick for LME 42 present in the data before the mapping? kind of - but different trend with clear smoothing after the pick. ALos lots of smoothing pre mapping with all countries. 
# effort_unmapped<-read_csv(file.path(yannick_dir, "Final_DataStudyFAO_AllGears_V5.csv")) # WARNING - was originally this Final_DataStudyFAO_V5.csv in local
# head(effort_unmapped)

effort_unmappedA<-effort_unmapped %>% 
  filter(SAUP == 36) %>% 
  group_by(Year, Gear) %>% 
  summarise(effort = sum(NomEffortActive, na.rm = TRUE)) %>% # look at number of vessels (NV) first - can also be NomEffort
  ungroup() %>% 
  mutate(origin = "unmapped")

all<-effort_unmappedA %>% 
  full_join(effort_mapped)

# WARNING - these are the same as here fishing country and gear are in common but then effort gets spread to different cells .... and it goes into different LMEs? 
ggplot(all, aes(x = Year, y = effort, color = origin))+
  geom_line()+
  theme_bw()+
  facet_wrap(~Gear)

# explore by country only - smoothing 
effort_unmapped2B<-effort_unmapped %>% 
  filter(Country == "Australia") %>% 
  group_by(Year) %>% 
  summarise(effort = sum(NV, na.rm = TRUE)) %>% 
  ungroup()

# Smoothing goes away as NV are allocated to the different gears based on a value that changes by year and given data on e.g. power 
ggplot(effort_unmapped2B, aes(x = Year, y = effort))+
  geom_line()

# most influential gears in determining the smooth pattern: 
# "Gillnets","Lines_Longlines","Seine_Danish_and_Other","Lines_Handlines_and_poles", "Others_Support"                  

effort_unmapped3<-effort_unmapped %>% 
  filter(Country == "Australia", !Gear %in% c("Gillnets","Lines_Longlines","Seine_Danish_and_Other","Lines_Handlines_and_poles", "Others_Support")) %>% 
  group_by(Year) %>% 
  dplyr::summarise(effort = sum(NV, na.rm=TRUE)) %>% 
  ungroup()

ggplot(effort_unmapped3, aes(x = Year, y = effort))+
  geom_line()

```

Explanation of key columns (For more info on the global effort file check Rousseau et al. in prep):  

*New*

* eez_country_name
* fao_area 
* LME
* SAUP
* Gear
* Sector 
* NomActive = effort in days at sea X kW (not creep-corrected)    
* EffActive = effort in days at sea X kW (creep-corrected)  
* NV
* P 
* GT 
* Year 

*Old* 

* FGroup = functional group   
* NomActive = effort in days at sea X kW (not creep-corrected)    
* EffActive = effort in days at sea X kW (creep-corrected)  
* NomActiveHours = effort in hours X kW (not creep-corrected) - *in old file version* 
* EffActiveHours = effort in hours X kW (creep-corrected) -  *in old file version* 
* Sectors = three main fishing sectors considered: I = Industrial, UP = un-powered artisanal, APW = artisanal powered    

NOTE: an updated version of this dataset will be provided. THis version will also include 'gear' and 'fishing country'.  

### Data manipulation

```{r global2, fig.width=10, fig.height=7}

# OLD - the new version already include LME etc... 

# # Load LME and FAO codes
# lme<-read.csv("/Users/camillan/Dropbox/DBPM_fishing_extension/Fishing data/LME_FAO_coding.csv")
# 
# # Calculate effort by LME, Year, Sector and Fleet
# effort_global<-effort %>% 
#   mutate(Fleet = FGroup) %>% # note - fleet should be based on fishing gear and target group  
#   group_by(LME, Year, Sector, Fleet) %>% 
#   summarise(NomActive = sum(NomActive, na.rm = TRUE)) %>%
#   ungroup() %>%
#   mutate(Year = as.numeric(Year)) 
# 
# # Add info from lme file - e.g. 'FAO' code or 'zone'
# lme<-lme %>% 
#   `colnames<-`(c("zone", "code","LME", "area")) %>% # rename columns as per effort data
#   filter(code == "LME") %>% # select only LME and do not consider FAO (can also do the opposite)
#   dplyr::select(-code, -area)
# 
# # Merge LME to effort data 
# effort_global<-effort_global %>% 
#   left_join(lme)

# Split effort by LME 
effort_list<-split(effort, effort$LME)

# Select South East Australia = LME 42
effort_42<-effort_list$'42' %>% 
  droplevels()

# are there fishing countries besides AU fishing in this LME that could explain the pick in 1990 not seen in regional effort data? see below 
unique(effort_42$SAUP)
# list the countries/SAUP codes: 
saup<-read_csv(file.path(yannick_dir, "SAUPcode_to_Country.csv")) 
head(saup)
# View(saup)

length(unique(effort_42$SAUP))

trial1<-effort_42 %>% 
  filter(SAUP %in% c(36, 157, 360, 764, 554)) %>% 
  group_by(Year, SAUP) %>% 
  summarise(effort = sum(NomActive, na.rm = TRUE)) %>% 
  ungroup()

ggplot(trial1, aes(x = Year, y = effort))+
  geom_line()+
  theme_bw()+
  facet_wrap(~SAUP)

# WARNING - lots of countries fishing in this LME but AU remains the one responsible for the pick in 1990

# are there fishing gears that explain this pick? 
trial2<-effort_42 %>% 
  group_by(Year, Sector, Gear) %>% 
  summarise(effort = sum(NomActive, na.rm = TRUE))

ggplot(trial2, aes(x = Year, y = effort, group = Sector, color = Sector))+
  geom_point(size=0.5)+
  geom_line()+
  theme_bw()+
  facet_wrap(~Gear)

# WARNING - yes midwater trawl explains the pick! 

# WARNING - could consider only industrial for LME 42 as modelers focus on industrial - should consider only Au and NZ as fishing countries?  
# effort_42<-effort_42 %>% filter(Sector == "Industrial")

# to compare with regional effort data
# WARNING - need to define fleets according to functional group too so keep FGroup
# WARNING - most of models are in D*Kw - can I do Global_effort_DayKW/P
effort_global<-effort_42 %>% 
  group_by(Gear, FGroup, Year) %>%
  summarise(Global_effort_DayKW = sum(NomActive, na.rm = TRUE),
            P = sum(P, na.rm = TRUE)) %>% 
  ungroup() %>% 
  mutate(Global_effort_Day = Global_effort_DayKW/P) %>% 
  select(-Global_effort_Day, -P) %>% # WARNING - not sure about the above so reverted back to DayKW
  `colnames<-`(c("Fleet", "FGroup", "Year","Global_effort_DayKW"))

# Aggregate effort by year
effort_global_aggregated<-effort_global %>%
  group_by(Year) %>%
  summarise(Global_effort_DayKW = sum(Global_effort_DayKW, na.rm = TRUE)) %>% 
  ungroup()

ggplot(effort_global_aggregated, aes(x = Year, y = Global_effort_DayKW))+
  geom_line()+
  theme_bw()

```

## Regional effort data

### Atlantis

```{r Atlantis, fig.width=10, fig.height=7}

# Two datasets available - see Beth's email "Effort and catch data from south east Australian models" 
models_dir <- "/rd/gem/private/users/yannickr/regional_models_LME42_effort"
effort_atlantis<-read.csv(file.path(models_dir, "1910-2003 SESSF hist info_effort.csv")) 
# WARNING - when using read_csv the dots on the column names disappear - just something to consider when using these 2 functions

# Calculate effort by fleet and year 
effort_atlantis<-effort_atlantis %>% 
  gather(key = "Type", value = "Effort",-Time)

# ggplot(effort_atlantis, aes(x = Time, y = Effort))+ 
#   geom_point(size = 0.5)+
#   geom_line()+
#   facet_wrap(~Type, scale = "free_y")+
#   theme_bw()

# WARNINGS - discussion with the modeller 
# WARNING - possible outlier 
effort_atlantis<-filter(effort_atlantis, Effort != 6.240000e+11)

# WARNING - some fleets only have 1 value in 2000, should they be filtered out? 
# out<-c("Abalone.divers", "Crab.trawl", "Demersal.line.for.deepwater.piscivorous.fish", "Lobster.pots", "Midwater.trawl.for.cephalopods", "Midwater.trawl.for.finfish", "Midwater.trawl.for.small.pelagics", "Lobster.pots", "Purse.seine.for.piscivorous.fish", "Purse.seine.for.small.pelagics", "Scallop.dredge","Jig.for.cephalopods","Traps.for.finfish")
# effort_atlantis<-effort_atlantis %>% filter(!Type %in% out)

# WARNING - are we double accounting for some of the effort - e.g. are Bottom.trawl.for.deepwater.demersal.fish;  Bottom.trawl.for.flathead; Bottom.trawl.for.orange.roughy somehow overlapping? discussed with Beth: not overlapping

# WARNING - purse.seine for large predator drives patters. Discussed with Beth: filter out
# effort_atlantis<-effort_atlantis %>% filter(Type != "Purse.seine.for.large.pelagics")

# Change columns names - more informative
colnames(effort_atlantis)<-c("Year", "Fleet", "Atl_effort_day")

# Plot final data 
ggplot(effort_atlantis, aes(x = Year, y = Atl_effort_day))+
  geom_point(size = 0.5)+
  geom_line()+
  theme_bw()+
  facet_wrap(~Fleet)

# Aggregate effort by year
effort_atlantis_aggregated<- effort_atlantis %>% 
  group_by(Year) %>% 
  summarise(Atl_effort_day = sum(Atl_effort_day, na.rm = TRUE))

```

### Mizer

```{r Mizer, fig.width=6, fig.height=4}

load(file.path(models_dir,"Mixed_fisheries_tradeoff.RData"))
effort_mizer = datValidationEffort

# WARNING - discuss with modeller
# WARNING - check that this is the most updated file
# WARNING - filter fleets that have not been considered 
effort_mizer<-filter(effort_mizer, metier != "Great Australian Bight - Trawl fishery")

# Change columns names - more informative
colnames(effort_mizer)<-c("Fleet","Year", "Mz_effort_opn")

ggplot(effort_mizer, aes(x = Year, y = Mz_effort_opn, group = Fleet))+
  geom_point(size = 0.5)+
  geom_line()+
  theme_bw()+
  facet_wrap(~Fleet)

# Aggregate effort by year
effort_mizer_aggregated <- effort_mizer %>% 
  group_by(Year) %>% 
  summarise(Mz_effort_opn = sum(Mz_effort_opn, na.rm= TRUE))

```

### EwE

```{r EwE, fig.width=6, fig.height=4}

# See Cathy's email "effort and catch data from south east Australian models" 25/02/2022 - new email with updated data in days at sea on 07/03/2022
effort_ewe<-read.csv(file.path(models_dir,"EBS Absolute effort values for Cami.csv"))
effort_ewe<-effort_ewe[-c(1,2),] # WARNING - what is pool code and type taht I deleted here? 

# WARNING - discuss with the modeller
# NOTES from cathy: scallops I would ignore. No Rec fishery real data.
# WARNING - squidJigs drives patters and rec fishing is not used. Discussed with Cathy and Beth: filter out 
effort_ewe<-effort_ewe %>% 
  gather("Fleet","Effort",-Name) %>% 
  mutate(Name = as.numeric(Name)) %>% 
  filter(!Fleet %in% c("Updated.Scallop")) 
  # filter(!Fleet %in% c("SquidEffort")) # see 

# need cleaner names... 
unique(effort_ewe$Fleet)
names<-data.frame(Fleet = unique(effort_ewe$Fleet), Fleet_new = c("Trawl", "NonTrawl", "LineVicNsw", "PrawnTrawlNsw", "Vic", "JigSquid", "PotTraps", "DanishSeine", "LineTuna"))

effort_ewe<-effort_ewe %>% 
  full_join(names) %>% 
  select(-Fleet) %>% 
  rename(Fleet = Fleet_new)

# Change columns names - more informative
effort_ewe<-effort_ewe[,c(1,3,2)]
colnames(effort_ewe)<-c("Year", "Fleet","Ewe_effort_day")

ggplot(effort_ewe, aes(x = Year, y = Ewe_effort_day))+ 
  geom_point(size = 0.5)+
  geom_line()+
  facet_wrap(~Fleet)+
  theme_bw()

# Aggregate effort by year
# WARNING - this is relative effort time series currently used to force the model. How do we merge these effort values? Here summed by year but data will be replaced.  
effort_ewe_aggregated<-effort_ewe %>% 
  group_by(Year) %>% 
  summarise(Ewe_effort_day = sum(Ewe_effort_day, na.rm = TRUE)) %>% 
  ungroup()

```

## Regional vs global

```{r all, fig.width=6, fig.height=4}

# Combine all effort data 
combined_effort<-effort_global_aggregated %>% 
  full_join(effort_atlantis_aggregated) %>% 
  full_join(effort_mizer_aggregated) %>% 
  full_join(effort_ewe_aggregated) %>% 
  dplyr::select(Year,Global_effort_DayKW, Atl_effort_day, Mz_effort_opn, Ewe_effort_day) %>% 
  gather(key = "Type", value = "Effort", -Year)

# Plot final data 
effort_plot<-ggplot(combined_effort, aes(x = Year, y = Effort))+
  geom_line(size = 1.2)+
  geom_vline(xintercept = 2005, linetype = "dashed")+
  facet_wrap(~Type, scales = "free_y")+
  theme_bw()
effort_plot

# # Save plot 
# setwd("/Users/camillan/Dropbox/DBPM_fishing_extension/Yannick_data/")
# pdf("Effort_LME42.pdf", width = 8, height=5)
# effort_plot
# dev.off()

# # PROBLEM - different effort units across datasets 
# # INDEPENDENTLY scaled from 0 to 1 to plot together. This is not a solution - could also do relative changes as but relative to different years?? 
# range01 <- function(x){(x-min(x, na.rm = TRUE))/(max(x, na.rm = TRUE)-min(x, na.rm = TRUE))}
# SEA_effort_global_aggregated$Global_effort_scaled<-range01(SEA_effort_global_aggregated$Global_effort_DayKW)
# effort_atlantis_aggregated$Atl_effort_scaled<-range01(effort_atlantis_aggregated$Atl_effort_day)
# effort_mizer_aggregated$Mz_effort_scaled<-range01(effort_mizer_aggregated$Mz_effort_opn)
# effort_ewe_aggregated$Ewe_effort_scaled<-range01(effort_ewe_aggregated$Ewe_effort_relative)
# 
# # Combine the 2 datasets again:
# combined_effort<-SEA_effort_global_aggregated %>% 
#   full_join(effort_atlantis_aggregated) %>% 
#   full_join(effort_mizer_aggregated) %>% 
#   full_join(effort_ewe_aggregated) %>% 
#   dplyr::select(Year,Global_effort_scaled, Atl_effort_scaled, Mz_effort_scaled, Ewe_effort_scaled) %>% 
#   gather(key = "Type", value = "Effort", -Year)
# 
# # Plot final data again
# effort_plot<-ggplot(combined_effort, aes(x = Year, y = Effort, group = Type, color = Type))+
#   geom_line(size = 0.7)+
#   geom_vline(xintercept = 2005, linetype ="dashed")+
#   theme_bw()
# effort_plot

# # Save plot
# pdf("Effort_LME42_scaled.pdf", width = 8, height=5)
# effort_plot
# dev.off()

```

#### Key points for discussion

1. Aggregated LME-level effort trends do not match across global and regional dataset:

  * Difficulties in comparing effort across datasets due to different units and little years overlap. much shorter time series for regional forcing data.      
  * Differences across models can be due to different 1) scales and 1) fleets/fisheries considered -> Exploration of effort by fleet might clarify some of this. The pick in 1990 in the global dataset is because of mid-water trawling - comparison by fleet is essential.

2. Atlantis and EwE show opposite trends between 1995 and 2005 -> Exploration of effort by fleet might clarify some of this.   
3. The global dataset seems to anticipate Atlantis in its increasing trend (1950-1990) and EwE in its decreasing trend (1990-2017). The pick in 1990 for the global dataset is due to mid-water trawling - comparison by fleet is essential.      
   
**NOTE**: if trends were matching -> **anomaly approach** could be used to force regional models with LME-level effort data from the global dataset. This would still require matching fleets across global and regional datasets - NEXT STEP. **statistical models** could be used to extract regional forcing using the relationship between regional and global effort at the fleet level. 

   
## Explore fishing effort by fleet 

```{r fleet, fig.width=10, fig.height=7}

# Manually create a file that stores and relates all fleets across the datasets
common_name<-unique(effort_global[,c(1,2)])
common_name$Fleet_global = paste(common_name$Fleet, common_name$FGroup, sep = "_")
common_name$Fleet_atl = NA 
common_name$Fleet_mz = NA 
common_name$Fleet_ewe = NA

common_name<-common_name[order(common_name$Fleet),]
# rownames(common_name)<-seq(1, nrow(common_name))

# # print and fill in .... 
# getwd()
# write.csv(common_name, "Data/fleet_mapping.csv")
# 
# # then upload and load in R the filled table fleet_mapping_filled.csv
# save.image(file = "Data/my_work_space.RData")

## WARNING units do not match when converting DayKW to Day by DayKW/P
original_effort_dir <- "/rd/gem/private/users/yannickr/effort_mapped_bycountry"

trial <- read.csv(file.path(original_effort_dir, "mapped_1950_APW_100.csv"))
head(trial)

#### Stat analysis ----
# Try with one net first to check th stat model and perhaps ask the modelers to fill in the table .... 

unique(effort_global$Fleet)
unique(effort_global$FGroup)
effort_global_Trawl<-effort_global %>% 
  filter(Fleet == "Trawl_Bottom", 
         FGroup %in% c("demersal>=90cm")) # only flathead gives a flat line for global
         # FGroup %in% c("flatfish<90cm", "demersal30-90cm", "demersal<30cm", "demersal>=90cm")) %>% # only flathead gives a flat line for global 
  # select(-Fleet, -FGroup) %>% 
  # group_by(Year) %>% 
  # summarise(Effort = sum(Global_effort_DayKW, na.rm = TRUE)) %>% # WARNING - effort is in different units fro global and atlantis ... what to do?? 
  # ungroup() %>% 
  # mutate(origin = "Global") 
  # rename(Effort = Global_effort_DayKW)

effort_atl_Trawl<-effort_atlantis %>% 
  filter(Fleet == "Bottom.trawl.for.flathead") %>% 
  select(c(-Fleet))
  # mutate(origin = "Atl") 
  # rename(Effort = Atl_effort_day)

effort_trawl<-effort_global_Trawl %>% 
  full_join(effort_atl_Trawl)

# ggplot(effort_trawl, aes(x = Year, y = Effort, group = origin, color = origin))+
#   geom_point()+
#   geom_line()

# View(effort_trawl)
# flathead - no relationship 
# demersal30-90cm, demersal<30cm negative relationship - meaningless 
# demersal>=90cm slightly increasing ... could try with this one ... 
ggplot(effort_trawl, aes(x = Atl_effort_day, y = Global_effort_DayKW))+
  geom_point()


# apply stat model 
# update.packages()
# install.packages("MuMIn")
# install.packages("GGally")
# library("MuMIn")
# library("GGally")


options(na.action = "na.fail")

fm1 <- lm(Atl_effort_day ~ Global_effort_DayKW, data = filter(effort_trawl, !is.na(Atl_effort_day)))
dd <- dredge(fm1)
# subset(dd, delta < 6)

# Visualize the model selection table:
# par(mar = c(3,5,6,4))
# plot(dd, labAsExpr = TRUE)

# Model average models with delta AICc < 4
model.avg(dd, subset = delta < 4)

# WARNING as above for BOATS
#or as a 95% confidence set:
model.avg(dd, subset = cumsum(weight) <= .95) # get averaged coefficients

#'Best' model
best<-summary(get.models(dd, 1)[[1]])
par(mfrow = c(2,2))
plot(get.models(dd, 1)[[1]])











#### SKYP as done in csv .... 

# Match Atlantis fleets 
sort(unique(effort_atlantis$Fleet))
unique(common_name$Fleet)
View(common_name)

# WARNING - need to discuss with Beth
common_name[, 4]<-"Abalone.divers" # Others_Unknown_demersalmollusc
common_name[,  4]<-"Bottom.trawl.for.benthopelagics" # Trawl_Bottom_benthopelagic>=90cm; Trawl_Bottom_benthopelagic30-90cm
common_name[, 4]<-"Bottom.trawl.for.cephalopods" # Trawl_Bottom_cephalopods
common_name[, 4]<-"Bottom.trawl.for.deepwater.demersal.fish" # Trawl_Bottom_demersal30-90cm, Trawl_Bottom_demersal<30cm, Trawl_Bottom_demersal>=90cm, Trawl_Bottom_rays<90cm, Trawl_Bottom_shark>=90cm, Trawl_Bottom_rays>=90cm, 
common_name[, 4]<-"Bottom.trawl.for.flathead" # Trawl_Bottom_flatfish<90cm
common_name[,  4]<-"Bottom.trawl.for.orange.roughy" # Trawl_Bottom_demersal30-90cm
common_name[, 4]<-"Crab.trawl" #
common_name[,  4]<-"Danish.seine"
common_name[, 4]<-"Demersal.finfish.gillnet"
common_name[, 4]<-"Demersal.line.for.deepwater.demersal.fish"
common_name[,  4]<-"Demersal.line.for.deepwater.piscivorous.fish" #
common_name[, 4]<-"Demersal.line.for.deepwater.shark"
common_name[, 4]<-"Demersal.line.for.shallow.water.finfish"
common_name[, 4]<-"Jig.for.cephalopods" # REPLACED 
common_name[, 4]<-"Lobster.pots" # REPLACED
common_name[, 4]<-"Midwater.trawl.for.cephalopods" # REPLACED
common_name[,  4]<-"Midwater.trawl.for.finfish" #
common_name[,  4]<-"Midwater.trawl.for.small.pelagics" #
common_name[, 4]<-"Pelagic.longline.for.large.pelagics"
common_name[, 4]<-"Prawn.trawl"
common_name[,  4]<-"Purse.seine.for.piscivorous.fish" # 
common_name[,  4]<-"Purse.seine.for.small.pelagics" #
common_name[, 4]<-"Scallop.dredge" # 
common_name[, 4]<-"Shark.gillnet"
common_name[, 4]<-"Traps.for.finfish" # 

# Match Mizer
common_name[25, 3]<-"GHAT - Southern Shark Gillnet"
common_name[6,  3]<-"South East Trawl Fishery - Danish Seine"
common_name[13, 3]<-"South East Trawl Fishery - Otter trawl deepSlope"
common_name[15, 3]<-"South East Trawl Fishery - Otter trawl shelf"
common_name[9,  3]<-"South East Trawl Fishery - Otter trawl upperSlope"

# Match EwE
# WARNING - need to discuss with Cathy
common_name[6,  4]<-"D.SeineEffort"
common_name[12, 4]<-"Line.effort"
common_name[22, 4]<-"NetNT.Effort" #
common_name[15, 4]<-"NSW.Effort"
common_name[14, 4]<-"ScallopVEffort"
common_name[16, 4]<-"TrapEffort"
common_name[9,  4]<-"Trawl.Effort"
common_name[18, 4]<-"TunaEffort"
common_name[21, 4]<-"VicEffort"

common_name$gearID<-common_name$Fleet_global

library("knitr")
knitr::kable(common_name[,1:4], format = "html")

# data for trends in effort by fleet
# 0 to 1 scaling function 
range01 <- function(x){(x-min(x, na.rm = TRUE))/(max(x, na.rm = TRUE)-min(x, na.rm = TRUE))}

# rearrange global effort data 
# WARNING - whether I ungroup() or not the effort_global df above makes a huge difference! 

effort_global2<-effort_global %>%
  select(Year, Fleet, Global_effort_DayKW) %>%
  mutate(Global_effort_scaled = range01(Global_effort_DayKW),
         data = "global") %>% # WARNING - check the scaling function
  rename(Fleet_global = Fleet) %>%
  full_join(common_name[,c(1,5)]) %>%
  filter(!is.na(Fleet_global))

ggplot(effort_global2, aes(x = Year, y = Global_effort_scaled, group = Fleet_global))+
  geom_point()+
  geom_line()+
  facet_wrap(~Fleet_global)

# rearrange Atl effort data
effort_atlantis2<-effort_atlantis %>%
  mutate(Atl_effort_scaled = range01(Atl_effort_day),
         data = "Atlantis") %>%
  rename(Fleet_atl = Fleet) %>%
  full_join(common_name[,c(2,5)])%>%
  filter(!is.na(Fleet_atl))

# rearrange Mz effort data
effort_mizer2<-effort_mizer %>%
  mutate(Mz_effort_scaled = range01(Mz_effort_opn),
         data = "Mizer") %>%
  rename(Fleet_mz = Fleet) %>%
  full_join(common_name[,c(3,5)]) %>%
  filter(!is.na(Fleet_mz))

# rearrange EwE effort data
effort_ewe2<-effort_ewe %>%
  mutate(Ewe_effort_scaled = range01(Ewe_effort_relative),
         data = "EwE")  %>%
  rename(Fleet_ewe = Fleet) %>%
  full_join(common_name[,c(4,5)])%>%
  filter(!is.na(Fleet_ewe))

# combine data
effort_all<-effort_global2 %>%
  full_join(effort_atlantis2) %>%
  full_join(effort_mizer2) %>%
  full_join(effort_ewe2)

effort_all<-effort_all %>%
  select(-c(Fleet_global,Global_effort_DayKW,Fleet_atl,Atl_effort_day,Fleet_mz,Mz_effort_opn,Fleet_ewe,Ewe_effort_relative)) %>%
  gather("key", "value", -c(Year, data, gearID)) %>% 
  ungroup()

ggplot(effort_all, aes(x = Year, y = value, color = data))+
  geom_point(size=0.5)+
  geom_line()+
  theme_bw()+
  facet_wrap(~gearID)

## prepare data for sankey plot
# Sum or mean effort across years and calculate proportional effort by fleet
effort_global_mean <- effort_global2 %>%
  group_by(Fleet_global, gearID) %>%
  summarise(Global_effort_DayKW = mean(Global_effort_DayKW, na.rm = TRUE)) %>% 
  ungroup() %>% # WARNING - this is needed otherwise the df is grouped by Fleet_global and the below operation gives wrong values
  mutate(Global_effort_prop = Global_effort_DayKW/sum(Global_effort_DayKW)) %>% 
  # rename(Fleet_global = Fleet) %>%
  select(-Global_effort_DayKW) # %>%
  # full_join(common_name[,c(1,5)]) %>%
  # filter(!is.na(Fleet_global))

effort_atlantis_mean<-effort_atlantis2 %>%
  group_by(Fleet_atl, gearID) %>%
  summarise(Atl_effort_day = mean(Atl_effort_day, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(Atl_effort_prop = Atl_effort_day/sum(Atl_effort_day)) %>%
  # rename(Fleet_atl = Fleet) %>%
  select(-Atl_effort_day) # %>%
  # full_join(common_name[,c(2,5)])%>%
  # filter(!is.na(Fleet_atl))

effort_mizer_mean<-effort_mizer2 %>%
  group_by(Fleet_mz, gearID) %>%
  summarise(Mz_effort_opn = mean(Mz_effort_opn, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(Mz_effort_prop = Mz_effort_opn/sum(Mz_effort_opn)) %>%
  # rename(Fleet_mz = Fleet) %>%
  select(-Mz_effort_opn) # %>%
  # full_join(common_name[,c(3,5)]) %>%
  # filter(!is.na(Fleet_mz))

effort_ewe_mean<-effort_ewe2 %>%
  group_by(Fleet_ewe, gearID) %>%
  summarise(Ewe_effort_relative = mean(Ewe_effort_relative, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(Ewe_effort_prop = Ewe_effort_relative/sum(Ewe_effort_relative))  %>%
  # rename(Fleet_ewe = Fleet) %>%
  select(-Ewe_effort_relative) #%>%
  # full_join(common_name[,c(4,5)])%>%
  # filter(!is.na(Fleet_ewe))

# combine data
effort_all_mean<-effort_global_mean %>%
  full_join(effort_atlantis_mean) %>%
  full_join(effort_mizer_mean) %>%
  full_join(effort_ewe_mean)

# View(effort_all_mean)

# define links global - atlantis
links1<-effort_all_mean %>%
  select(Fleet_global, Fleet_atl, Atl_effort_prop) %>% # WARNING - this does not take into consideration how effort is spread across the global dataset... So not sure how useful this is
  filter(!is.na(Atl_effort_prop))

colnames(links1)<-c("source", "target", "value")

# From these flows we need to create a node data frame: it lists every entities involved in the flow
nodes <- data.frame(
  name=c(as.character(links1$source),
  as.character(links1$target)) %>% unique()
)

# With networkD3, connection must be provided using id, not using real name like in the links dataframe.. So we need to reformat it.
links1$IDsource <- match(links1$source, nodes$name)-1
links1$IDtarget <- match(links1$target, nodes$name)-1

# Make the Network
p <- sankeyNetwork(Links = links1, Nodes = nodes,
              Source = "IDsource", Target = "IDtarget",
              Value = "value", NodeID = "name",
              sinksRight=FALSE)
p

# define links Atlantis - EwE
links2<-effort_all_mean %>%
  select(Fleet_atl,Fleet_ewe ,Ewe_effort_prop) %>%
  filter(!is.na(Ewe_effort_prop))

colnames(links2)<-c("source", "target", "value")

# combine links global - atlantis - mizer
links_all<-rbind(links1[,c(1,2,3)], links2) # delete ID columns calculated above

nodes <- data.frame(
  name=c(as.character(links_all$source),
  as.character(links_all$target)) %>% unique()
)

# With networkD3, connection must be provided using id, not using real name like in the links dataframe.. So we need to reformat it.
links_all$IDsource <- match(links_all$source, nodes$name)-1
links_all$IDtarget <- match(links_all$target, nodes$name)-1

# Make the Network
p <- sankeyNetwork(Links = links_all, Nodes = nodes,
              Source = "IDsource", Target = "IDtarget",
              Value = "value", NodeID = "name",
              sinksRight=FALSE)
p

# define links Atl - Mizer
# WARNING - atl - mz works because there are no NA between the two (each mz fleet is inked to an Atl fleet). But EwE - mz do not work at the moment for the opposite reason. Anyways if there are NA between 2 models, the links are messed up... needs more thoughts 

# View(effort_all_mean)

links3<-effort_all_mean %>%
  select(Fleet_atl,Fleet_mz ,Mz_effort_prop) %>%
  filter(!is.na(Mz_effort_prop))

colnames(links3)<-c("source", "target", "value")

# combine links global - atlantis - mizer
links_all<-rbind(links1[,c(1,2,3)], links2, links3) # delete ID columns calcualted above

nodes <- data.frame(
  name=c(as.character(links_all$source),
  as.character(links_all$target)) %>% unique()
)

# With networkD3, connection must be provided using id, not using real name like in the links dataframe.. So we need to reformat it.
links_all$IDsource <- match(links_all$source, nodes$name)-1
links_all$IDtarget <- match(links_all$target, nodes$name)-1

# Make the Network
p <- sankeyNetwork(Links = links_all, Nodes = nodes,
              Source = "IDsource", Target = "IDtarget",
              Value = "value", NodeID = "name",
              sinksRight=FALSE)
p

```
